ML Estimation of Matrix--$t$ Distributions with Expectation Maximization

### 2017-02-02 paper.tex
Work on wording of introduction

\section{Introduction}
The matrix--$t$ distribution is a generalization of the multivariate $t$ distribution to a matrix variate situation. Similar as other $t$ distributions, the matrix--$t$ distribution is a scale mixture of a particular class of multivariate normal models, the matrix normal distribution, except the mixture parameter is a matrix rather than a scalar \citep{gupta1999matrix}. Due to this, the distribution shares its importance in robust estimation and has been applied to many predictive tasks such as spatial interpolation on the prediction of pollution concentration \citep{KIBRIA2006785} and recommendation systems \citep{NIPS2007_3203}.

These matrix variate models, including matrix normal distributions \citep{2013arXiv1309.6609G}, give better predictive results than the traditional PCA methods as the correlation structure among different observations can be captured in one of the scale matrix parameter and hence do not require the $i.i.d$ assumption. In these inferential problems, it is necessary to obtain the parameters of the underlying model. For matrix--$t$ distributions, the parameters consist of a degrees of freedom parameter, a matrix location parameter, and two scale matrix parameters, which describe a set of given matrix--valued observations for further inference.

A common approach on estimating the parameters is through the method of maximum likelihood (ML). However, optimizing the likelihood directly is a formidable task as the gradient equations are unwieldy slow to compute and solve. For $t$ distributions, it is well-known that one can exploit the mixture property and apply expectation maximization (EM) technique to optimize the likelihood function \citep{10.2307/24305551}. The same technique can be also applied to the matrix variate situation by viewing it as a matrix-normal distribution with inverse Wishart distributed scale parameter. When missing data presents, imputation of the missing values can be considered as a prediction problem and the relevant work for matrix--$t$ distributions can be found in \cite{NIPS2007_3203}.

This paper is organized as follows. In section \ref{sec:matrixt} we give a brief review on matrix--$t$ distributions. Notations and preliminary matrix calculus results needed to present the derivation are presented sequentially in the section. EM formulae together with the relevant derivation and implementation details are provided in section \ref{sec:EM} and a brief discussion is given in section \ref{sec:discussion}.

>>>>

\section{Introduction}
The matrix--$t$ distribution is a generalization of the multivariate
$t$ distribution to matrix variate domains. Similar to other $t$
distributions, the matrix--$t$ distribution is a scale mixture of
a particular class of multivariate normal models: In particular, it is
a mixture of a matrix normal distribution, with a matrix-valued scale
parameter \citep{gupta1999matrix}.  Because of its mixture nature, the
matrix--$t$ distribution can be applied to building predictive models
with applications in tasks such as  spatial interpolation (for
example, to predict the concentration of pollution
\citep{KIBRIA2006785}) and recommendation systems
\citep{NIPS2007_3203}. Matrix models can be advantageous compared to
more traditional PCA methods because the correlation structure among
different observations can be captured in the scale matrix parameters,
and hence do not require the $i.i.d$ assumption
\citep{2013arXiv1309.6609G}. In such inferential problems, it is
required to obtain the parameters of the underlying model, given a set
of observed data samples. For matrix--$t$ distributions, the
parameters consist of a degrees of freedom parameter, a matrix
location parameter, and two scale matrix parameters.

The method of maximum likelihood (ML) is a common approach to estimate
the parameters of the distribution. However, for matrix--$t$
distributions, optimizing the likelihood directly is impractical since
the gradient equations are unwieldy. Instead, we can exploit the fact
that general for $t$ distributions, it is well-known that one can use
the mixture property and apply the expectation maximization (EM)
technique to optimize the likelihood function
\citep{10.2307/24305551}. EM can also be  applied to the particular
case of a matrix variate domains, by viewing the matrix--$t$
distribution as a mixture.

The related problem of determining a sample from a matrix--$t$
distribution given some observed entries of the matrix and parameters
of the model was studied in \cite{NIPS2007_3203}.

This paper is organized as follows: In section \ref{sec:matrixt} we
give a brief review on matrix--$t$ distributions, fix the notation and
give some matrix calculus results. In section \ref{sec:EM}, we derive
the EM formulae and provide relevant implementation details.  A brief
discussion is given in section \ref{sec:discussion}.
